---
title: "DS241 Final Proeject - DC Bikeshare "
output: html_notebook
author: "Alexis Alexander, Gianna Capoccia, Morgan Eaton, and Mateo Sarabia"
date: "09 December 2023"
---

## Background Information

The purpose of our analysis involves exploring potential methods of increasing ridership for the DC Captial Bikeshare program. This motivation in focusing on ridership aligns with the DC Bikeshare's main mission in transforming the Washingtion Metropolitan Area and residing communities by providing an affordable bicycle transit system. The goals outlined in this targeted reformation includes decreasing traffic congestion, promoting health and wellness, decreasing air pollution, and expanding transportation options. The DC Biekshare program is owned by several jurisdictions however, is mainly managed by the DC local government.  

More information pertaining to the bikeshare program can be found by following this link: https://capitalbikeshare.com/

## Business Understanding 

Following the methods outline in the CRISP-DM Methodology, we initially focused on understanding the various business aspects to our problem before analyzing any data. We determine our business objectives by researching the background of the DC Bikeshare program, as seen above. Next, we determine our data mining goals and identified a overall question we aim to solve for our analysis.

This question pertains to the following: How do we make it safer for people to ride bikes as a means of increasing ridership?


## Installing Packages 

Below are the packages should be installed:

```{r}
install.packages("tidyverse")
install.packages("janitor")
install.packages("lubridate")
install.packages("here")
install.packages("mapview")
install.packages("gbfs")
install.packages("sf")
install.packages("tmap")
install.packages("tidycensus")
```

Baseline packages for our analysis: 

```{r}
library(tidyverse)
library(janitor)
library(lubridate) # because we will probably see some dates
library(here) # more easily access files in your project
library(mapview)
library(gbfs)
library(sf) # working with simple features - geospatial
library(tmap)
library(tidycensus)
```

## Accessing and Loading the Databases 

We are initially looking at the following three data sets:

1. The DC Bikeshare System Data for September 2023: https://capitalbikeshare.com/system-data
2. The DC Protected vs Unprotected Bikelane Dataset: https://opendata.dc.gov/datasets/bicycle-lanes/explore
3. The DC Crimes and Accidents Dataset: https://opendata.dc.gov/datasets/crashes-in-dc/explore

```{r}
bikes_september <- read_csv(here("data_raw","202309-capitalbikeshare-tripdata.csv"))%>% clean_names() 
```

```{r}
bikes_lanes = st_read(here("data_raw", "Bicycle_Lanes.geojson")) %>% clean_names()
```

```{r}
bikes_crashes = st_read(here("data_raw", "Crashes_in_DC.geojson")) %>% clean_names()
```
## Comparsion of Bicycle Related Crashes to Major Bike Station 

Comparing Bike Lanes and Crashes 

Filtering the crash data to work with a smaller time frame, looking specifically at the month of September 2023. First, we selected variables that we are interested in looking at from the total data set. We then filtered the data to only look at accidents relating the bicycles. This was determined by setting a requirement that there must be at least 1 bicycle involved in the accident. 

For the bike lane data set, we also reduced the number of variables to only the ones we were interested in analyzing. We also noticed that the "Ward" was common variable that was included in both the crash and bike lane data set. This will be useful when we later inter-join the two data sets. 

```{r}
simplified_bike_crashes = bikes_crashes %>%
  select(fromdate, address, latitude, longitude, xcoord, ycoord, ward, majorinjuries_bicyclist, minorinjuries_bicyclist, fatal_bicyclist, total_bicycles, total_vehicles, bicyclistsimpaired, speeding_involved) %>%
   filter(total_bicycles >= "1") 

simplified_bike_lanes = bikes_lanes %>% 
  select(routename, bikelane_parkinglane_adjacent, bikelane_throughlane_adjacent, bikelane_pocketlane_adjacent, bikelane_contraflow, bikelane_conventional, bikelane_dual_protected, bikelane_dual_buffered, bikelane_protected, bikelane_buffered, streetname, streettype, totalbikelanes, totalbikelanewidth, ward_id, geometry, objectid)
```


Now that we have a simplified data set, we are interested in the crash data that occur in September 2023. First, we will separate the date and time from each other, as they are initially recorded in the same column together. 

```{r}
simplified_bike_crashes$date <- as.Date(simplified_bike_crashes$fromdate)
simplified_bike_crashes$time <- format(simplified_bike_crashes$fromdate, "%H: %M: %S")
simplified_bike_crashes$month <- format(simplified_bike_crashes$date, "%m")
simplified_bike_crashes$year <- format(simplified_bike_crashes$date, "%y")
```

Next, we will filter the data to look at the data from September 2023 

```{r}
Sept_simplified_bike_crashes = simplified_bike_crashes %>%
  filter(month == "09" & year =="23")
```

Reducing our data set is essential to making it easier for us to analyze our data. We will later look at a longer time period. 

We will now inter_join the bike lane and crash data based on the Ward specified. However, the Ward columns of the two dataset does not match for each observation recorded. The crash data uses "Ward #" while, the lane data just reports the #. First, we will make sure that each observation is recorded similarly. We also removed any NA from the data as well. 

```{r}
Sept_simplified_bike_crashes_ward = Sept_simplified_bike_crashes %>%
  separate(ward, c(NA, "Ward"))
```

Now, that the observations are recorded in the same way. We will attempt to inter_join the two data sets. Later, we recognized that our data set is a sf class thus, we need to use st_join to join the two data sets. 


```{r}
bike_crashes_lanes = Sept_simplified_bike_crashes_ward %>%
  st_join(simplified_bike_lanes)
class(bike_crashes_lanes)
```

We will create an initial map of our data. First, we are setting the tmap mode to interactive viewing 

```{r}
tmap_mode("view")
```

%% Mapping of Our Initial Data 

Plotting the crash data spatially 

```{r}
Sept_simplified_bike_crashes_ward %>%
  tm_shape()+
  tm_dots()

```

```{r}
simplified_bike_lanes %>%
  tm_shape()+
  tm_dots()
```



```{r}
bike_crashes_lanes %>%
  tm_shape()+
  tm_dots()
```


## Issue with Joining the Crash and Bike Lane Together - Try Using the CSV files instead

We will come back to the mapping data at a later date. 

We noticed that after using the st_join, the bike lane data did not join correctly with the crash data. To re-evaluate our data, we decided to load the data as csv instead of using the geosjeon file. 

```{r}
csv_bike_lane <- read_csv(here("data_raw","Bicycle_Lanes.csv"))%>% clean_names() 
csv_bike_crashes <- read_csv(here("data_raw","Crashes_in_DC.csv"))
```


```{r}
csv_simplified_bike_crashes = csv_bike_crashes %>%
  select(FROMDATE, ADDRESS, LATITUDE, LONGITUDE, XCOORD, YCOORD, WARD, MAJORINJURIES_BICYCLIST, MINORINJURIES_BICYCLIST, FATAL_BICYCLIST, TOTAL_BICYCLES, TOTAL_VEHICLES, BICYCLISTSIMPAIRED, SPEEDING_INVOLVED) %>%
   filter(TOTAL_BICYCLES >= "1") 

csv_simplified_bike_lanes = csv_bike_lane %>% 
  select(routename, bikelane_parkinglane_adjacent, bikelane_throughlane_adjacent, bikelane_pocketlane_adjacent, bikelane_contraflow, bikelane_conventional, bikelane_dual_protected, bikelane_dual_buffered, bikelane_protected, bikelane_buffered, streetname, streettype, totalbikelanes, totalbikelanewidth, ward_id, objectid)
```


```{r}
csv_simplified_bike_crashes$date <- as.Date(csv_simplified_bike_crashes$FROMDATE)
# csv_simplified_bike_crashes$time <- format(csv_simplified_bike_crashes$FROMDATE, "%H: %M: %S")
csv_simplified_bike_crashes$month <- format(csv_simplified_bike_crashes$date, "%m")
csv_simplified_bike_crashes$year <- format(csv_simplified_bike_crashes$date, "%y")
```


```{r}
csv_Sept_simplified_bike_crashes = simplified_bike_crashes %>%
  filter(month == "09" & year =="23")
```


```{r}
csv_Sept_simplified_bike_crashes_ward = Sept_simplified_bike_crashes %>%
  separate(ward, c(NA, "Ward"))
```

Want to join the two data sets based on the route name or the street name that is given. 

```{r}
csv_crash_address = csv_Sept_simplified_bike_crashes_ward %>%
  separate(address, c("address number", "streetname","street type","quadrant"))
```
Choosing to delete rows that did not separate correctly as shown in the error above 

```{r}
csv_removed_crash_address = csv_crash_address[-c(2,8,10,15,37), ]
```

Simplify the bike lane data so that instead of defining the lane as a collection of points, we instead want it to count only once in our data. The bike  lane data can be split to differentiate between how the bike lane flows with traffic vs. whether there are any physical barriers or distinct separations in place. Right now, we will solely look at the physical barriers and the code below filters our data as such. 

```{r}
csv_bike_lanes_reduced = csv_simplified_bike_lanes %>%
  count(streetname, bikelane_dual_protected, bikelane_dual_buffered, bikelane_protected, bikelane_buffered) %>%
  drop_na(streetname) %>%
  group_by(streetname) 

```


States the number of bike lanes associated with each street 
```{r}
test2 = csv_simplified_bike_lanes %>%
  count(streetname)
```


```{r}
test3 = csv_simplified_bike_lanes %>%
  select(streetname, bikelane_dual_protected)

 # test3$bikelane_dual_protected <- replace_na("None")
```


Testing code to obtain the number of bike lanes for each type that is associated with each street

```{r}
test = csv_simplified_bike_lanes %>%
  count(streetname, bikelane_conventional) %>%
  drop_na(bikelane_conventional) %>%
  group_by(streetname) %>%
  summarise(bikelane_conventional_total = sum(n))
```





Left_join the crash and bike lane data set based on the street name 
```{r}
csv_crash_lanes = csv_removed_crash_address %>%
  left_join(csv_simplified_bike_lanes, "streetname")
glimpse(csv_crash_lanes)
```




## Finding the Number of Accidents Associated with Each Bike Lane  


## Creating a Variable for Classifying the Type of Accident




Calculating the number of accidents that occurred for each bike lane 

```{r}

```

Calculate the a sum crashes that occurs along the bike lane line. 
The bike lane is a geometric line, while each crash is recorded as a geometric point. 




