---
title: "DS241 Final Proeject - DC Bikeshare "
output: html_notebook
author: "Alexis Alexander, Gianna Capoccia, Morgan Eaton, and Mateo Sarabia"
date: "09 December 2023"
---

## Background Information

The purpose of our analysis involves exploring potential methods of increasing ridership for the DC Captial Bikeshare program. This motivation in focusing on ridership aligns with the DC Bikeshare's main mission in transforming the Washingtion Metropolitan Area and residing communities by providing an affordable bicycle transit system. The goals outlined in this targeted reformation includes decreasing traffic congestion, promoting health and wellness, decreasing air pollution, and expanding transportation options. The DC Biekshare program is owned by several jurisdictions however, is mainly managed by the DC local government.  

More information pertaining to the bikeshare program can be found by following this link: https://capitalbikeshare.com/

## Business Understanding 

Following the methods outline in the CRISP-DM Methodology, we initially focused on understanding the various business aspects to our problem before analyzing any data. We determine our business objectives by researching the background of the DC Bikeshare program, as seen above. Next, we determine our data mining goals and identified a overall question we aim to solve for our analysis.

This question pertains to the following: How do we make it safer for people to ride bikes as a means of increasing ridership?


## Installing Packages 

Below are the packages should be installed:

```{r}
install.packages("tidyverse")
install.packages("janitor")
install.packages("lubridate")
install.packages("here")
install.packages("mapview")
install.packages("gbfs")
install.packages("sf")
install.packages("tmap")
install.packages("tidycensus")

```

Baseline packages for our analysis: 

```{r}
library(tidyverse)
library(janitor)
library(lubridate) # because we will probably see some dates
library(here) # more easily access files in your project
library(mapview)
library(gbfs)
library(sf) # working with simple features - geospatial
library(tmap)
library(tidycensus)

```

## Accessing and Loading the Databases 

We are initially looking at the following three data sets:

1. The DC Bikeshare System Data for September 2023: https://capitalbikeshare.com/system-data
2. The DC Protected vs Unprotected Bikelane Dataset: https://opendata.dc.gov/datasets/bicycle-lanes/explore
3. The DC Crimes and Accidents Dataset: https://opendata.dc.gov/datasets/crashes-in-dc/explore

```{r}
library(readr)
bikes_september <- read_csv("~/Documents/data sci/ds241_f23/data_raw/202309-capitalbikeshare-tripdata.csv") %>% clean_names()


library(readr)
bikes_lanes <- read_csv("~/Documents/data sci/ds241_f23/data_raw/Bicycle_Lanes-2.csv") %>% clean_names()

 library(readr)
bikes_crashes <- read_csv("~/Documents/data sci/ds241_f23/data_raw/Crashes_in_DC-2.csv") %>% clean_names()
```

##  Create a visualization that compares the type of injury with the type of bike lane.
## More protected lane results in a less severe injury? 
## Can also look at bike lane width***

```{r}
                        ## filter out any accidents before jan 1st 2019
bikes_crashes2 = bikes_crashes |> filter(reportdate >= "2019/01/01 05:00:00+00")

                ## major bicyclist injuries
df1 = bikes_crashes2 |> filter(majorinjuries_bicyclist>=1)

                ## seperate the desired column
  df1a <- df1[ -c(1:22,24:58) ]
  
                ## minor bicyclist injuries
df2 = bikes_crashes2 |> filter(minorinjuries_bicyclist>=1)

                ## seperate the desired column
  df2a <- df2[ -c(1:23,25:58) ]
  
  
                ## fatal bicyclist injuries
df3 = bikes_crashes2 |> filter(fatal_bicyclist>=1)

                ## seperate the desired column
  df3a <- df3[ -c(1:25,27:58) ]
```


```{r}
## contraflow bikelanes"
df4 = bikes_lanesgeo |> filter(bikelane_contraflow != "NA")

 ## seperate the desired column
  df4a <- df4[ -c(1:7,9:26) ]

## conventional bikelanes 
df5 = bikes_lanesgeo |> filter(bikelane_conventional != "NA")

 ## seperate the desired column
  df5a <- df5[ -c(1:8,10:26) ]

## dual protected bikelanes
df6 = bikes_lanesgeo |> filter(bikelane_dual_protected != "NA")

 ## seperate the desired column
  df6a <- df6[ -c(1:9,11:26) ]

## dual buffered bikelanes 
df7 = bikes_lanesgeo |> filter(bikelane_dual_buffered != "NA")

 ## seperate the desired column
  df7a <- df7[ -c(1:10,12:26) ]

## protected bikelanes
df8 = bikes_lanesgeo |> filter(bikelane_protected != "NA")

 ## seperate the desired column
  df8a <- df8[ -c(1:11,13:26) ]

## buffered bikelanes 
df9 = bikes_lanesgeo |> filter(bikelane_buffered != "NA")

 ## seperate the desired column
  df9a <- df9[ -c(1:12,14:26) ]

```


```{r}
dfa = cross_join(df1a,df4a)
class(dfa)
ggplot(data = dfa, mapping = aes(x = majorinjuries_bicyclist, y = bikelane_contraflow)) + geom_point()
```

```{r}
tmap_mode("view")
```

```{r}
bikes_lanesgeo = st_read("~/Documents/data sci/ds241_f23/data_raw/Bicycle_Lanes.geojson") %>% clean_names()

bikes_crashesgeo = st_read("~/Documents/data sci/ds241_f23/data_raw/Crashes_in_DC.geojson") %>% clean_names()
```

```{r}
tmap_mode("view")
```

```{r}

bikes_lanesgeo2 %>%
  tm_shape()+
  tm_lines(c("bikelane_contraflow"))
```







```{r}
## bikelane contraflow
df4 %>%
  tm_shape()+
  tm_lines("red")

```


```{r}
#bikelane conventional
df5 %>%
  tm_shape()+
  tm_lines("red")
```

```{r}
#bikelane dual protected
df6 %>%
  tm_shape()+
  tm_lines("red")

```

```{r}
#bikelane dual buffered
df7 %>%
  tm_shape()+
  tm_lines("red")
```

```{r}
#bikelane protected
df8 %>%
  tm_shape()+
  tm_lines("red")
```

```{r}
#bikelane buffered
df9 %>%
  tm_shape()+
  tm_lines("red")
```





